---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

Two data types: **vectors** and **nodes**.
Vectors can be:
- Atomic: all elements of the same type
- List
- NULL: generic zero-length vector

Vector attributes:
- Dimension
- Class: for OOP


**Atomic vectors**
- Logical (boolean)
- double (decimal, scientific (12.3E4), or hex)
- integer (L suffix)
- character (strings)
- complex
- raw (bytes)

```{r}
?Quotes #how to represent various characters
```

```{r}
x <- 3 #create a simple vector
```

```{r}
x
```

```{r}
str(x) #print the structure of x, 'num' = numeric
```

```{r}
class(x)
```

```{r}
typeof(x)
```

```{r}
length(x)
```

```{r}
y <- 3L #integer
class(y)
typeof(y)
length(y)
str(y)
```

```{r}
z <- x>0 #create a logical vector
class(z)
typeof(z)
length(z)
str(z)
```

```{r}
w <- 'three' #character
class(w)
typeof(w)
length(z)
str(z)
```

Scalar do not exist: they are one-element vectors

```{r}
?c #concatenate function
```

```{r}
x <- c(1)
str(x)
```

```{r}
x <- c(3, 5, 6, 3, 1, 2, -10)
str(x)
length(x)
```

Vectors cannot be extended - you need to create a longer vector and then copy the older one.

```{r}
0:10 #sequence (endpoints are included)
```

```{r}
15:5 #backward sequence
```

```{r}
seq(-2, 3, 0.5) #as in np.arange
```

```{r}
seq(from=0.04, to=0.15, length=6) #as np.linspace
```

```{r}
rep(9, 5) #constant vectors
```

```{r}
rep(1:4, 2) #replicate sequence (np.tile)
```

```{r}
rep(1:4, each=2, times=3) #np.repeat + np.tile
```

```{r}
rep(1:4, 1:4) #np.repeat with a np.arange
```

```{r}
rep(c("cat", "dog", "mouse"), c(2,3,2))
```

## Indeterminate forms

```{r}
4/0
```

```{r}
-15/0
```

```{r}
exp(-Inf) #infinity are propagated!
```

```{r}
0/Inf
```

```{r}
(0:3)^Inf
```

```{r}
0/0
Inf - Inf
Inf/Inf
```

```{r}
#Test infinites
x <- -4.5
is.finite(x)

is.infinite(c(-4.5, 0/0, exp(Inf)))

is.nan(c(-4.5, 0/0, exp(Inf)))
```

## Unknown values

```{r}
#Missing data are represented as NA
NA > 0; 2.7*NA; ! NA #normally all operations on NA result in NA
```

```{r}
#except for operations with a unique result
NA^0
NA |  TRUE
NA & FALSE
```

```{r}
y <- c(4, NA, -8)

y == NA #does not work to check the NA
y == "NA" #NA is not a string!

#should use:
is.na(y)
```

```{r}
#Remove NA
y[! is.na(y)]
```

```{r}
#Some operations can "skip" NAs
v1 <- c(1:7, NA, 2:8, NA, 3:9)
str(v1)
```

```{r}
mean(y) #NA "ruin" the operation
```

```{r}
mean(v1, na.rm=TRUE) #remove the NA before computing the mean
```

```{r}
#Get indices of NAs
seq(along=v1) #create sequence of all indices

seq(along=v1)[is.na(v1)] #select only the ones that correspond to NAs
```

```{r}
#or simpler way:
which(is.na(v1))
```

```{r}
#Replace NAs with a value (e.g. 0)
v1[is.na(v1)] <- 0
v1
```

```{r}
#Equivalently
v1[which(is.na(v1))] <- 0
v1
```

```{r}
ifelse(is.na(v1), 0, v1) #vectorized ifelse (np.where)
```

```{r}
probe <- 1:5
```

```{r}
probe
```

```{r}
probe[-1] #remove the first
```

```{r}
#Create a function to remove the smallest two values and larges two values
trim <- function(x) {
    sort(x)[-c(1,2, length(x)-1, length(x))] #remove with the -
}
```

```{r}
probe <- c(4, 7, 6, 5, 6, 7)
probe
```

```{r}
trim(probe)
```

## Vectors attributes

```{r}
counts <- c(25, 12, 7, 4, 6, 2, 1, 0, 2)
attr(counts, "nx") <- "count1"
attr(counts, "ny") <- "events"
```

```{r}
attr(counts, "nx")
```

```{r}
attributes(counts)
```

Achtung! Many operations *destroy* attributes.

```{r}
x <- c(a = 1, b = 2, c = 3)
str(x) #give names to the elements
```

```{r}
x <- 1:3
names(x) <- c("a", "b", "c")
str(x)
```

```{r}
x <- setNames(1:3, c("a", "b", "c"))
str(x)
```

```{r}
names(x)
```

```{r}
#Remove
unname(x)

#or
names(x) <- NULL
str(x)
```

```{r}
#for example, set names to represent histogram labels
counts <- c(25, 12, 7, 4, 6, 2, 1, 0, 2)
names(counts) <- 0:(length(counts)-1)
str(counts)
```

```{r}
hist(counts) #does not use names
```

```{r}
barplot(counts) #use names
```

```{r}
v1 <- c(1:20)
dim(v1) <- c(4,5) #np.reshape to 2d array
class(v1)
str(v1)
```

```{r}
v1
```

```{r}
v1 <- c(1:20)
dim(v1) <- c(2, 5, 2) #np.reshape to 3d array

str(v1)
```

```{r}
v1
```

```{r}
#Sum all elements less than 5
x <- 0:10
sum(x[x<5])
```

## S3 atomic vectors
S3 is the basic object system in R. 
- factor vectors (categorical data)
- Date vectors (day resolution)
- POSIXct/POSIXlt vectors (second resolution)
- difftime vectors (durations)


```{r}
x <- factor(c("a", "b", "c", "d")) #factor
str(x)
typeof(x)
attributes(x)
```

```{r}
coord <- factor(c("Est", "West", "Est", "North"), levels = c("North", "Est", "South", "West"))
table(coord) #count occurrences
```

```{r}
#ordered factors (when order matters)
#e.g. security level
grade <- ordered(c("high", "low", "medium", "medium"), levels = c("low", "medium", "high"))
str(grade)
```

```{r}
summary(grade)
```

```{r}
barplot(table(grade))
```

```{r}
#read.csv(), data.frame(), etc. convert character vectors to factors. Can disable with stringsAsFactors = FALSE
```

```{r}
today <- Sys.Date(); today #date
```

```{r}
typeof(today)
class(today)
```

```{r}
yesterday <- as.Date("2020-03-14")
yesterday
```

```{r}
delta <- today - yesterday
str(delta)
class(delta) #difftime
```

```{r}
#internal representation is accessed with unclass
unclass(today) #= number of days since 1970-01-01
```

```{r}
#POSIX = Portable Operating System Interface, a standard from UNIX
#ct = calendar time (time_t in C)
#lt = local time    (struct_tm type in C)
```

```{r}
now_ct <- as.POSIXct(Sys.time(), tzone="CET")
now_ct
```

```{r}
r20bday_ct <- as.POSIXct("2020-02-29 12:00", tzone="CET")
now_ct - r20bday_ct
```

```{r}
structure(now_ct, tzone="Europe/Rome")
structure(now_ct, tzone="Europe/Moscow") #edits only the formatting, not the internal representation
structure(now_ct, tzone="Asia/Chongqing")
```

```{r}
#time durations
one_week <- as.difftime(1, units="weeks")
attributes(one_week)
```

```{r}
today <- Sys.time()
next_week <- today + one_week
structure(next_week, tzone="Europe/Rome")
```

```{r}
forty_min <- as.difftime(40, units="mins")
```

## Unique and duplicated

```{r}
names <- c("John", "John", "Jim", "Anna", "Beatrix", "Anna")
table(names)
```

```{r}
unique(names)
```

```{r}
duplicated(names) #mark elements that have appeared before
```

```{r}
names[!duplicated(names)] #same as unique
```

```{r}
setA <- c("a", "b", "c", "d", "e")
setB <- c("d", "e", "f", "g")

union(setA, setB)
```

```{r}
intersect(setA, setB)
```

```{r}
setdiff(setA, setB)
```

```{r}
setdiff(setB, setA)
```

```{r}
setequal(setA, setA)
setequal(setA, setB)
```

```{r}
setA %in% setB #marks elements in the intersection
```

```{r}
9E-4/0.05
```

```{r}

```
