---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r}
x <- c(1,2,3)
y <- x
y[3] <- 4

str(x)
str(y)
```

Variables are references. When copying x in y, we are not copying the data, but we have two references to the same object. If then $y$ changes, R creates a copy laso of the data.

To inspect the internal data structure and memory allocation we can use the package lobstr.

```{r}
install.packages("lobstr")
```

```{r}
lobstr::obj_addr(x) #we can use the package name as a namespace, with "::" to access its methods
```

```{r}
help(package=lobstr) #Access documentation
```

```{r}
library(lobstr) #loads the package, so that we do not need to use its namespace
```

```{r}
#So, let's redo the above experiment and inspect it with lobstr
```

```{r}
a <- c(1,2,3)
b <- a

obj_addr(a); obj_addr(b) #same address
```

```{r}
b[1] <- 2

obj_addr(a); obj_addr(b) #address of b changes
```

## Lists

```{r}
l1 <- list( 1:3, "list element", c(TRUE, FALSE, FALSE), c(3.5, 4, 6.2, -1.75))
typeof(l1)
```

Lists contain several different objects, as a list of references to them. R has a garbage collector, and so automatically deallocates memory when the last reference to an object goes out of scope.

```{r}
ref(l1) #inspect the list: gives the address to each of its objects
```

## Matrices

```{r}
X <- matrix (c(1,0,0,0,1,0,0,0,1) , nrow =3) ; X
```

```{r}
class(X)
attributes(X)
str(X)
```

```{r}
vct <- c(1,2,3,4,4,3,2,1)
V <- matrix ( vct , byrow =T , nrow =2) #equivalent to vct.reshape((4,2))
V
```

```{r}
V <- matrix ( vct , byrow =F , nrow =2)
V
```

```{r}
X <- matrix(rpois(n=20, lambda=1.5), nrow=4) #Fill matrix with samples from Poisson distribution
X
```

```{r}
X[3,3]
```

```{r}
X[4,] #row 4
```

```{r}
X[,5] #column 5
```

```{r}
rowSums(X)
```

```{r}
rowMeans(X)
```

```{r}
#Adding rows and columns

vct <- matrix(c(1,0,2,5,1,1,3,1,3,1,0,2,1,0,2,1), byrow=T, nrow=4)
vct
```

```{r}
vct <- rbind(vct, apply(vct, 2, mean)) #rbind combines "by row", i.e. merges the vectors returned by the apply function as rows
```

```{r}
vct <- cbind(vct, apply(vct, 1, var))
```

```{r}
#apply(matrix, axis, functin): iterates over vectors on axis (1=rows, 2=columns) of that matrix
```

```{r}
colnames(vct) <- c(1:4, "variance") #Name the fields
rownames(vct) <- c(1:4, "mean")
#Note that 1:4 are converted to characters (names are always characters)
```

```{r}
vct
```

```{r}
(Y <- matrix(rbinom(20, 9, 0.45), nrow=4)) #adding brackets around prints the result
```

```{r}
apply(Y, MARGIN=2, FUN=sum) #sum the values in columns
```

```{r}
apply(Y, 1, function(x) x^2+x) #can use lambda functions
```

```{r}
sapply(12:14, seq) #applies function to a vector, generates three sequences: 1:12, 1:13 and 1:14 (seq(12), seq(13), seq(14))
```

```{r}
#Set random seed
set.seed(2019)
runif(3) #uniform U(0,1) distribution
```

```{r}
#at any moment we can save the current random seed
current.seed <- .Random.seed
```

```{r}
runif(3)
```

```{r}
runif(3)
```

```{r}
current.seed -> .Random.seed #resets the seed
runif(5) #we get the same sequence as before!
```

```{r}
y <- c(8,3,5,7,6,6,8,9,2,3,9,4,10,4,11)

sample(y) #permutates the elements (sampling WITHOUT replacement)
```

```{r}
sample(y, 3) #samples only 3 elements
```

```{r}
sample (y , replace =T ) #with replacement
```

```{r}
x <- 1:10
sample(x[x>8])

```

```{r}
sample(x[x>9]) #x[x>9] has only one element (10). In this case, sample generates a sequence 1:10 and samples from it instead

#to avoid this behaviour, use resample:
library(gdata) #need this package for resample function

resample(x[x>8])
resample(x[x>9])
```

```{r}
?resample
```

From the resample docs: resample differs from the S/R sample function in resample always considers x to be a vector of elements to select from, while sample treats a vector of length one as a special case and samples from 1:x. Otherwise, the functions have identical behavior.


## R Subsetting

```{r}
x <- c(2.1, 4, 6.7, 1.75)
```

```{r}
x[c(1,3)] #get first and third element
```

```{r}
x[c(1,1,3,3)] #duplicate indices work
```

```{r}
x[sort(x)] #doubles are rounded to ints for indexing
```

```{r}
#negative integers exclude elements
x[-c(1,3)]
```

```{r}
#cannot use positive and negative at the same time
x[c(1,-3)]
```

```{r}
x[c(T, T, F, T)] #logical vectors work as masks
```

```{r}
x[c(TRUE, FALSE)] #if the selection vector is too short, it is repeated (as c(T,F,T,F))
```

```{r}
x[0] #returns a 0-length vector
```

```{r}
#Named vectors are just like dictionaries
#some useful constant vectors: LETTERS (capital letters), letters (lowercase), month.abb, month.name
```

```{r}
y <- setNames(x, LETTERS[1:length(x)])
y
```

```{r}
y["A"] #access through the character indices
```

```{r}
y[c('A', 'A', 'D')]
```

```{r}
#Subsetting with factors uses the underlying integer vector, not the character!
y[factor("B")] #returns the first element, because "B" in the factor is stored as a 1
#In general, avoid using factors for subsetting
```

```{r}
#outer = cartesian product
outer(1:3, 1:3)
```

```{r}
#a function can be specified to edit the generated elements
#paste = converts arguments to characters, and concatenates them (with a separator if specified)

v <- outer(1:5, 1:5, FUN="paste", sep=",")
v
```

```{r}
v[seq(3, 23, 5)] #we can access elements in the matrix as if it were a "flattened vector"
#(matrices are internally stored as sequential data)
```

```{r}
#Preserve original dimension when subsetting: DROP=FALSE
(S <- matrix(1:6, nrow=2))
```

```{r}
S[1,] #normally returns a vector
```

```{r}
S[1, , drop=FALSE] #returns a (1,3) matrix instead
```

```{r}
#Lists are like "trains". [[]] gets the content of a vagon, [] just the vagon itself
xl <- list(1:3, "one", c(T,F,F)) #3 "vagons"
```

```{r}
xl[1] #first vagon, returned as a LIST
```

```{r}
xl[[1]] #first vagon, returned as VECTOR (=content of the "vagon")
```

```{r}
xl[1:2] #subset the list
```

```{r}
xl[[1:3]] # = xl[[1]][[2]]
```

```{r}
xl[[1]][[3]]
```

## Loops

```{r}
#Define a factorial function
fact1 <- function(x)
{
    f <- 1
    if (x < 2) return (1)
    for (i in 2:x) {
        f <- f*i
    }
    return (f)
}
```

```{r}
sapply(1:5, fact1)
```

```{r}
#with a while loop instead
fact2 <- function(x) {
    f <- 1; t <- x
    while (t > 1) {
        f <- f*t
        t <- t-1
    }
    return(f)
}
```

```{r}
sapply(1:5, fact2)
```

```{r}
fac3 <- function(x) {
    f <- 1; t <- x
    repeat { #= while(1)
        if (t<2) break
        f <- f*t
        t <- t-1
    }
    return(f)
}
sapply(1:5, fac3)
```

```{r}
cumprod(1:5) #cumulative product
```

```{r}
#Better way: use already defined (vectorized) functions!
fac4 <- function(x) max(cumprod(1:x))
```

```{r}
sapply(1:5, fac4)
```

```{r}
#Also now there is the factorial function:
sapply(1:5, factorial)
```

```{r}
#AVOID LOOPS!
```

```{r}
#Es. use ifelse vectorized function
y <- log(rpois(20,1.5))
y
```

```{r}
mean(y)
```

```{r}
(y <- ifelse(y<0, NA, y)) #negative values are set to NA

mean(y, na.rm=TRUE) #so that they can be avoided by functions
```

```{r}
x <- runif(50000000)
str(x)
```

```{r}
head(x)
tail(x)
```

```{r}
system.time(max(x))
```

```{r}
pc <- proc.time()
cmax <- x[1]
for (i in 2:length(x)) { if (x[i]>cmax) cmax <- x[i]}
proc.time()-pc
```

```{r}
test1 <- function(n) {
    y <- 1:n #optimized
}

test2 <- function(n) {
    y <- numeric(n) #empty array
    for (i in 1:n)  #initialize with for loop
        y[i] <- i
}

test3 <- function(n) { 
    y <- NULL
    for (i in 1:n) #even worse
        y <- c(y,i) #creates a new array every time!
}
```

```{r}
system.time(test1(10000000))
system.time(test2(10000000))
#system.time(test3(10000000)) too long
```

```{r}

```
