---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# Exercise_r01
**Name**: Francesco Manzali <br/>
**ID**: 1234428

```{r}
sessionInfo()
```

# Exercise 1

```{r}
install.packages('units') #to add units
```

```{r}
library(units)
```

```{r}
loch <- paste0('Loch ', c('Ness', 'Lomond', 'Morar', 'Tay', 'Awe', 'Maree',
                          'Ericht', 'Lochy', 'Rannoch', 'Shiel', 'Katrine', 'Arkaig', 'Shin'))
volume <- c(7.45, 2.6, 2.3, 1.6, 1.2, 1.09, 1.08, 1.07, 0.97, 0.79, 0.77, 0.75, 0.35)
units(volume) <- "km^3"
area <- c(56, 71, 27, 26.4, 39, 28.6, 18.6, 16, 19, 19.5, 12.4, 16, 22.5)
units(area) <- "km^2"
length <- c(39, 36, 18.8, 23, 41, 20, 23, 16, 15.7, 28, 12.9, 19.3, 27.8)
units(length) <- "km"
max_depth <- c(230, 190, 310, 150, 94, 114, 156, 162, 134, 128, 151, 109, 49)
units(max_depth) <- "m"
mean_depth <- c(132, 37, 87, 60.6, 32, 38, 57.6, 70, 51, 40, 43.4, 46.5, 15.5)
units(mean_depth) <- "m"
```

```{r}
#0. Build dataframe from vectors
(scottish.lakes = data.frame(loch, volume, area, length, max_depth, mean_depth))
```

```{r}
#1. Evalaute the highest and lowest volume and area lake
volume_area = scottish.lakes[,c(2,3)]
sprintf('Maximum')
apply(volume_area, 2, max)
sprintf('Minimum')
apply(volume_area, 2, min)
```

```{r}
#2. Order the frame with respect to the area and determine the two largest area lakes
(area_sorted = scottish.lakes[order(area, decreasing=TRUE),])
```

```{r}
#Two lakes with highest area
area_sorted[1:2,]
```

```{r}
#3. By summing up the areas occupied by the lakes, determine the area of Scotland covered by water
sprintf('Area of Scotland covered by water is %d km^2', sum(scottish.lakes$area))
```

# Exercise 2

```{r}
install.packages(c('DAAG', 'tibble'), type='source')
```

```{r}
#Load libraries
library(tibble)
library(DAAG)
```

```{r}
#Get information on the package content and on the ais data frame
?tibble
```

```{r}
library(help='DAAG') #DAAG contains several datasets
```

```{r}
?ais #measurements of blood characteristics for various athletes
```

```{r}
#Create a tibble from the ais data.frame
(dct <- tibble(ais)) 
```

```{r}
#1. Group data by gender and sport. Produce a barplot with the table adding a legend
gender_sport = table(dct$sex,dct$sport)

barplot(gender_sport, beside=TRUE, legend = rownames(gender_sport), las=2, ylab='Count')
```

```{r}
#2. Do columns hold missing values?
any(is.na(dct)) #No missing values
```

```{r}
#3. Make boxplots of the main blood variables (rcc, wcc, hc, hg) for different kind of sports
old_par <- par()

par(mfrow=c(2,2))

boxplot(rcc ~ sport, data = dct, las=2,
        xlab = 'Sport', ylab = 'Red blood cell count', col='firebrick3', medcol='white')

boxplot(wcc ~ sport, data = dct, las=2,
        xlab = 'Sport', ylab = 'White blood cell count', col='ghostwhite', medcol='dodgerblue2')

boxplot(hc ~ sport, data = dct, las=2,
        xlab = 'Sport', ylab = 'Hematocrit', col='darkorchid3', medcol='white')

boxplot(hg ~ sport, data = dct, las=2,
        xlab = 'Sport', ylab = 'Hemaglobin', col='deeppink3', medcol='white')
```

```{r}
#4. Scatter plots for correlations
plot(dct$rcc, dct$wcc, col = ifelse(dct$sex == 'f', "deeppink3", "dodgerblue2"), pch = ifelse(dct$sex == 'f', 16, 15),
     xlab = 'Red blood cell count', ylab = 'White blood cell count')
legend(6, 14, legend=c("Female", "Male"),
       col=c("deeppink3", "dodgerblue2"), pch=c(16,15), cex=0.8)

plot(dct$rcc, dct$hc, col = ifelse(dct$sex == 'f', "deeppink3", "dodgerblue2"), pch = ifelse(dct$sex == 'f', 16, 15),
     xlab = 'Red blood cell count', ylab = 'Hematocrit')
legend(6, 60, legend=c("Female", "Male"),
       col=c("deeppink3", "dodgerblue2"), pch=c(16,15), cex=0.8)

plot(dct$rcc, dct$hg, col = ifelse(dct$sex == 'f', "deeppink3", "dodgerblue2"), pch = ifelse(dct$sex == 'f', 16, 15),
     xlab = 'Red blood cell count', ylab = 'Hemaglobing concentration')
legend(6, 14, legend=c("Female", "Male"),
       col=c("deeppink3", "dodgerblue2"), pch=c(16,15), cex=0.8)
```

# Exercise 3

```{r}
needed_packages <- c('lubridate', 'readxl', 'curl')
```

```{r}
already_installed <- needed_packages %in% installed.packages()
for (pack in needed_packages[!already_installed]) {
    message(paste("To be installed:", pack, sep=" "))
    install.packages(pack)
}
```

```{r}
library(lubridate)
library(readxl)
library(curl)
```

```{r}
url <- "https://www.ecdc.europa.eu/sites/default/files/documents/"
fname <- "COVID-19-geographic-disbtribution-worldwide-"
date <- lubridate::today() - 1
ext = ".xlsx"
target <- paste(url, fname, date, ext, sep="")
message("target: ", target)
tmp_file <- tempfile("data", tempdir(), fileext=ext)
tmp <- curl::curl_download(target, destfile=tmp_file)
```

```{r}
covid <- readxl::read_xlsx(tmp_file)
```

**Data structure**
- `dateRep`: date of the report
- `day`, `month`, `year`: reformatted `dateRep`
- `cases`, `deaths`: confirmed cases and deaths on that day
- `countriesAndTerritories`: location (extended name)
- `geoId`: geographical identifier
- `countryterritoryCode`: country identifier
- `popData2018`: total population in that region (according to World Bank 2018 database)

```{r}
#1. Examine data
covid
```

```{r}
#2. Create a sub-tibble containing only the last day (hint: apply a selection on dateRep column)
(last_day = covid[covid$dateRep == date,])

#Produce a table with all countries with number of deaths or number of new cases greather than 200
most_affected_countries = last_day[last_day$cases > 200 | last_day$deaths > 200, 'countriesAndTerritories']
```

```{r}
most_affected_countries
```

```{r}
#3. Select the top 10 countries, in terms of cases, and plot the total number of cases as a function of time.
#Plot the total number of deaths as a function of time.
#In order to compare the different curves, normalize the first date-time plot to the same t0 value.

top10countries = as.list(last_day[order(last_day$cases, decreasing=TRUE),][1:10,'countriesAndTerritories'])[[1]]
#Top 10 countries in terms of cases during last day
```

```{r}
top10countries
```

```{r}
covid_selected = covid[covid$countriesAndTerritories %in% top10countries,] #Select only most affected countries
covid_ordered = covid_selected[order(covid_selected$dateRep),] #order by date

#for each country, add its cumulative sum of cases in the column "cumulative_cases"
covid_ordered$cumulative_cases <- ave(covid_ordered$cases, covid_ordered$countriesAndTerritories, FUN=cumsum)

#same for deaths
covid_ordered$cumulative_deaths <- ave(covid_ordered$deaths, covid_ordered$countriesAndTerritories, FUN=cumsum)
```

```{r}
library(ggplot2)

date_plot <- function(p) {
    p + geom_line(aes(colour = factor(countriesAndTerritories))) + xlab('Date') + ylab('Total number of cases') +
    xlim(as.POSIXlt("2020-03-01"), as.POSIXlt("2020-04-14")) + scale_color_discrete(name="Country")
}

p <- ggplot(covid_ordered, aes(dateRep, cumulative_cases))
p <- date_plot(p)
print(p)

p2 <- ggplot(covid_ordered, aes(dateRep, cumulative_deaths))
p2 <- date_plot(p2) + ylab("Total number of deaths")
print(p2)
```

```{r}
threshold = 1000
thresholded = subset(covid_ordered, cumulative_cases > threshold) #start counting days from when cumulative_cases surpasses threshold
```

```{r}
start_dates = as.list(by(thresholded$dateRep, thresholded$countriesAndTerritories, FUN=min))
```

```{r}
starts = unlist(start_dates[covid_ordered$countriesAndTerritories], use.names=FALSE) #starting date for each row. Convert list => numerical array
attributes(starts) <- attributes(covid_ordered$dateRep) #convert numerical array => posixct
```

```{r}
covid_ordered$days_past_infection = difftime(covid_ordered$dateRep, starts, units="days")
```

```{r}
p3 <- ggplot(subset(covid_ordered, days_past_infection > 0), aes(days_past_infection, cumulative_cases)) + 
      geom_line(aes(colour = factor(countriesAndTerritories))) + scale_color_discrete(name="Country") + xlab('Days past first 1000 cases') +
      ylab('Total number of cases')
print(p3)

p4 <- ggplot(subset(covid_ordered, days_past_infection > 0), aes(days_past_infection, cumulative_deaths)) + 
      geom_line(aes(colour = factor(countriesAndTerritories))) + scale_color_discrete(name="Country") + xlab('Days past first 1000 cases') +
      ylab('Total number of deaths')
print(p4)
```
