---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r}
options(repr.plot.width=8, repr.plot.height=5) #set a common width/height for plots
```

## Exercise 1
A publishing company has recently launched a new journal. In order to determine how effective it is in reaching its possible audience, a market survey company selects a random sample of people from a possible target audience and interviews them. Out of $150$ interviewed people, $29$ have read the last issue of the journal.

a) What kind of distribution would you assume for $y$, the number of people that have seen the last issue of the journal?

```{r}

```

b) Assuming a uniform prior, what is the posterior distribution for $y$?

```{r}

```

c) Plot both posterior and likelihood distributions functions

```{r}

```

## Exercise 2

```{r}

```

```{r}

```
