---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# Installation
**Anaconda**
- `conda install -c r r-essentials` (existing environment)
- `conda create -n environment_name -c r r-essentials` (new environment)

**Docker**
```bash
docker pull r-base #install an image
docker images -a #list images
open xquartz #open xserver
ifconfig | fgrep 192 #get local IP address
xhost + <address> #config local ip address ?```

`docker run -it --rm -v "$PWD":/mnt -e DISPLAY=address:0 --name Rint r-base  #run docker interactive`
- -it = interactive session
- -rm remove container on exit
- -v "$PWD":/mnt virtual folder /mnt linked to local folder PWD
- -e DISPLAY=address:0 configurate graphical display
- --name Rint name for the docker session
- r-base which image to run
    

`docker run --rm -d -P --name RJup -v "$PWD":/home/jovyan/work jupyter/r-notebook`

- -d = detached session (run in background)
- -P returns a port to access the docker (needed for jupyter)
- --name RJup
- -v "$PWD":/home/jovyan/work #workfolder of the notebook
- jupyter/r-notebook #name of the image
    
```docker ps -a #see the ports mapping
docker logs --tail 4 #see the token for authentication```
    
#open jupyter in a browser and use the token


## Get Help

```{r}
?sqrt #help for a certain package
```

```{r}
apropos("lm") #Related packages
```

```{r}
example(sqrt) #run example code
```

```{r}
help.search("data input") #fuzzy search in the docs
```

```{r}
find("read.table") #find a package
```

```{r}
library(help=base) #list contents of "base" package
```

```{r}
pkg <- installed.packages() #store the list of all the installed packages in a variable "pkg"
```

```{r}
pkg #it appears as a table
```

```{r}
df_pkg <- data.frame(pkg) #organize the tabular data in a DataFrame structure
str(df_pkg) #show the structure
```

```{r}
names(df_pkg) #column names
```

```{r}
df_pkg[1,2] #Access row 1, column 2
```

```{r}
df_pkg[1,] #Just first row, all columns
```

```{r}
df_pkg[1,1:3] #first row, first 3 columns
```

```{r}
df_pkg[1:10,1:3] #first 10 rows, first 3 columns
```

```{r}
df_pkg[,1:3] #all rows, first 3 columns
```

**Install new packages**
- from CRAN (official, verified, *stable* packages)
- from GitHub repos (developer versions)
- other repositories (e.g. BioConductor)

```{r}
#es. data.table: https://cran.r-project.org/web/packages/data.table/index.html, or on https://github.com/Rdatatable/data.table
```

## Manage objects

```{r}
objects() #or ls(), list all the objects created within the session
```

```{r}
search()
```

```{r}
install.packages("data.table") #install the data.table
```

```{r}
#Normally new packages are not loaded:
search()
```

```{r}
library("data.table") #Load the new package
search()
```

## R as calculator

```{r}
log(34/5.5)
```

```{r}
log(34.7) + sqrt(12) -
 25 / 7 * 46^3 #can split 
```

```{r}

```
