---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r}
x <- c(1,2,3)
y <- x
y[3] <- 4

str(x)
str(y)
```

Variables are references. When copying x in y, we are not copying the data, but we have two references to the same object. If then $y$ changes, R creates a copy laso of the data.

To inspect the internal data structure and memory allocation we can use the package lobstr.

```{r}
install.packages("lobstr")
```

```{r}
lobstr::obj_addr(x) #we can use the package name as a namespace, with "::" to access its methods
```

```{r}
help(package=lobstr) #Access documentation
```

```{r}
library(lobstr) #loads the package, so that we do not need to use its namespace
```

```{r}
#So, let's redo the above experiment and inspect it with lobstr
```

```{r}
a <- c(1,2,3)
b <- a

obj_addr(a); obj_addr(b) #same address
```

```{r}
b[1] <- 2

obj_addr(a); obj_addr(b) #address of b changes
```

## Lists

```{r}
l1 <- list( 1:3, "list element", c(TRUE, FALSE, FALSE), c(3.5, 4, 6.2, -1.75))
typeof(l1)
```

Lists contain several different objects, as a list of references to them. R has a garbage collector, and so automatically deallocates memory when the last reference to an object goes out of scope.

```{r}
ref(l1) #inspect the list: gives the address to each of its objects
```

## Matrices

```{r}
X <- matrix (c(1,0,0,0,1,0,0,0,1) , nrow =3) ; X
```

```{r}
class(X)
attributes(X)
str(X)
```

```{r}
vct <- c(1,2,3,4,4,3,2,1)
V <- matrix ( vct , byrow =T , nrow =2) #equivalent to vct.reshape((4,2))
V
```

```{r}
V <- matrix ( vct , byrow =F , nrow =2)
V
```

```{r}
X <- matrix(rpois(n=20, lambda=1.5), nrow=4) #Fill matrix with samples from Poisson distribution
X
```

```{r}
X[3,3]
```

```{r}
X[4,] #row 4
```

```{r}
X[,5] #column 5
```

```{r}
rowSums(X)
```

```{r}
rowMeans(X)
```

```{r}
#Adding rows and columns

vct <- matrix(c(1,0,2,5,1,1,3,1,3,1,0,2,1,0,2,1), byrow=T, nrow=4)
vct
```

```{r}
vct <- rbind(vct, apply(vct, 2, mean)) #rbind combines "by row", i.e. merges the vectors returned by the apply function as rows
```

```{r}
vct <- cbind(vct, apply(vct, 1, var))
```

```{r}
#apply(matrix, axis, functin): iterates over vectors on axis (1=rows, 2=columns) of that matrix
```

```{r}
colnames(vct) <- c(1:4, "variance") #Name the fields
rownames(vct) <- c(1:4, "mean")
#Note that 1:4 are converted to characters (names are always characters)
```

```{r}
vct
```

```{r}
()
```
