---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# Exercise 1
The number of particles emitted by a radioactive source during a fixed interval of time ($\Delta t = 10$s) follows a Poisson distribution on the parameter $\mu$. The number of particles observed during consecutive time intervals is $4$, $1$, $3$, $1$ and $3$.

(a) Suppose a uniform prior distribution for the parameter $\mu$.
- determine and draw the posterior distribution for $\mu$, given the data

```{r}
data   <- c(4, 1, 3, 1, 3)
deltat <- 10

likelihood_a <- function(mu, data) {
    res <- 1
    
    for (datum in data) {
        res <- res * dpois(datum, lambda=mu)
    }
    
    return(res)
}

posterior_a <- function(mu, data) {
    #Posterior is just proportional to the likelihood, so we just need to normalize it
    
    norm <- integrate(likelihood_a, lower=0, upper=20, data=data)$value
    return(likelihood_a(mu, data) / norm)
}
```

```{r}
plot(0, 0, "n", xlim=c(0, 10), ylim=c(0,.6), xlab=expression(mu), ylab="Posterior")
curve(posterior_a(x, data), from=0, to=10, add=TRUE) #Numerically, by normalizing the likelihood (as the prior is a uniform)
curve(dgamma(x, sum(data) + 1, rate=length(data)), add=TRUE, col='red', lwd=2, lty=2)
legend("topright", legend=c("Numerical", "Analytical"), lwd=c(1,2), lty=c(1,2), col=c("black", "red"), y.intersp=1.5)
#Analytically, the posterior for a poisson process with uniform prior is a Gamma pdf with alpha = sum(x_j) + 1 and lambda=n,
#where {x_j}_{j=1,...,n} are the n Poisson observations
```

```{r}

```
